---
title: "Final_Project_Script"
author: "Caleb Milford"
date: "Spring 25"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Installing and loading initial packages

```{r}
# install.packages("tidyverse")
# install.packages("rmarkdown")
# install.packages("terra")

library(tidyverse)
library(rmarkdown)
library(terra)

```

## My raw data

The data that I am using "Combines_Frog_Project_Data.csv" is the culmination of all the herpetological surveys that have been done over the past three years from the Bullfrog crew that is working throughout southeast Arizona from the Bauder lab. During these surveys they record all herps that are encountered during every survey as well as all other relevant data at the time of encounter. This is everything from the surveyors names, species encountered, location in UTM coordinates, site level covariates like vegetation and much more. My primary goal throughout this script will be to clean and midify this data set so that it is usable for my first thesis chapter.

My thesis chapter will only be looking at two species specifically and doing some occupancy modeling for both of them individually. The two species are the Sonoran mud turtle (*Kinosternon sonoriense*) and the Sonora desert toad (*Incilius alvarius*) and I am primarily interested in how site level and landscape level covariates are affecting the presence/ absence of these two species. Considering this initial data set has a lot of extra information in it. My first goals will be to pull out the appropriate info so I can continue analysis.

**Loading in raw data**

```{r}
#getwd()
all_herp_data <- read_csv("../data_raw/Combined_Frog_Project_Data.csv")
all_herp_data
```

### Week 3: Intro to data frames

Goal is to parse down the data frame and remove unnecessary columns or info that is not relevant to my project

```{r}
#Removing completely unnesesary columns 
all_herp_data2 <- all_herp_data %>% 
  select(-surveyors, -Adult, -Adults_removed, -Juv, -Juv_removed, -Larvae,
         -Larvae_removed, -Egg_mass, -County, -Final_Start_NAD83_East,
         -Final_Start_NAD83_North) %>% 
  rename("Taxon" = taxon) %>% 
  arrange(Date)
         
#Just to more easiliy visualize what I am looking at without so much outside info 
all_herp_data_obrev <- all_herp_data2 %>% 
  select(Date, Start_time, Site_name, Species, No_Obs)
all_herp_data_obrev
```

### Week 6: Cleaning untidy data

```{r}
#Replacing unclear notation or 0's with NA's

all_herp_data2 <- all_herp_data2 %>% 
  mutate(Air_temp_150cm_C = na_if(Air_temp_150cm_C, "n/r"), 
         Water_tmp_5cm_C = na_if(Water_tmp_5cm_C, "n/r"),
         Relative_humidity = na_if(Relative_humidity, "n/r"),
         Floating_species = na_if(Floating_species, "0"),
         Submerged_species = na_if(Submerged_species, "0"), 
         Emergent_species = na_if(Emergent_species, "0"))
  
#seperating the date column into day month year columns 
all_herp_data2 <- all_herp_data2 %>% 
  separate(Date, c("Year", "Month", "Day"), sep = "-")

#Changing the data frame to one row per survey instead of one row per species found 
all_herp_data3 <- all_herp_data2 %>%
  group_by(Year, Month, Day, Site_name, Start_time) %>%
  summarise(
    Species = paste(unique(Species), collapse = ", "),
    .groups = "drop"
  )
#This worked however it did not maintain the orignal data frames number of columns. It only gave back the columns I grouped by. Need to come back and figure out how to have each row represent 1 survey while maintaining all origial columns 

#Summarize detections by month/year for each species 
#This data will help us to determine site closure when lookig at occupancy 
KISO_monthly_det <- all_herp_data3 %>% 
  filter(Species == "KISO") %>% 
  group_by(Year, Month) %>% 
  summarise(Detections = n(), .groups = "drop")
KISO_monthly_det

write_csv(KISO_monthly_det, "../outputs/KISO_monthly_det.csv")

BUAL_monthly_det <- all_herp_data3 %>% 
  filter(Species == "BUAL") %>% 
  group_by(Year, Month) %>% 
  summarise(Detections = n(), .groups = "drop")

write_csv(BUAL_monthly_det, "../outputs/BUAL_monthly_det.csv")

```

### Week 5: Data Visualization

```{r}
all_herp_data2 %>% 
  filter(Species == "KISO") %>% 
  ggplot(aes(x = Air_temp_150cm_C, y = Water_tmp_5cm_C))+
  geom_point()+
  labs(x = "Air Temp (C)", y = "Water Temp (C)")+
  theme_bw()

all_herp_data2 %>% 
  filter(Species == "BUAL") %>% 
  ggplot(aes(x = Air_temp_150cm_C, y = Water_tmp_5cm_C))+
  geom_point()+
  labs(x = "Air Temp (C)", y = "Water Temp (C)")+
  theme_bw()


#This part still needs a lot of work dont judge me 
```
